---
title: "Results"
author: "Hans Kohls"
date: "23/05/2015"
output: html_document
---

# Introduction

As part of the Coursera Data Science Specialization this report summarizes the activities undertaken to complete the project for the Machine Learning course. 

As part of the excercise, data from the Human Activity Recognigtion project was used to determine 5 classes of excercises with dumbells. The dataset is available at http://groupware.les.inf.puc-rio.br/har.

# Data Exploration

The data that was imported consisted of numerous empty columns which were also not reflected in the dataset supplied for validaton in the project. So a total of 50 measurements was used as features to identify the activity class.

```{r eval=FALSE}

data_columns_row=c(
    "roll_belt", "pitch_belt", "yaw_belt", "total_accel_belt",
    "gyros_belt_x", "gyros_belt_y", "gyros_belt_z",
    "accel_belt_x", "accel_belt_y", "accel_belt_z",
    "magnet_belt_x", "magnet_belt_y", "magnet_belt_z",
    "roll_arm", "pitch_arm", "yaw_arm", 
    "gyros_arm_x", "gyros_arm_y", "gyros_arm_z",
    "accel_arm_x", "accel_arm_y", "accel_arm_z",
    "magnet_arm_x", "magnet_arm_y", "magnet_arm_z",
    "roll_dumbbell", "pitch_dumbbell", "yaw_dumbbell",
    "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z",
    "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z",
    "magnet_dumbbell_x", "magnet_dumbbell_y", "magnet_dumbbell_z",
    "roll_forearm", "pitch_forearm", "yaw_forearm",
    "total_accel_forearm", 
    "gyros_forearm_x", "gyros_forearm_y", "gyros_forearm_z", 
    "accel_forearm_x", "accel_forearm_y", "accel_forearm_z",
    "magnet_forearm_x", "magnet_forearm_y", "magnet_forearm_z",
    # Activity Class
    "classe")

```

## Pre-Processing

Based on a number of early models I identified that various training methods available in the caret package could not handle more than between 1 and 10% of the full data set as a training set on my environment. I also noticed strong correlations between some of the features. 

To improve the outcome, I applied centering and scaling as preprocessing methods. I also tested a reduction of the degrees of freedom for each record by applying primary component analysis, but this was dismissed in the end. 

```{r eval=FALSE}
preProc <- preProcess(training[,1:50], method=c('center', 'scale'))
trainPP <- predict(preProc,training[,1:50])
testPP <- predict(preProc,testing[,1:50])
projectPP <- predict(preProc,project)

```

In the final model, my system could handle training data at a rate of 30% of the full data set. So data was generated accordingly.

```{r eval=FALSE}
inTrain <- createDataPartition(y=data_by_row$classe, p=0.3, list=F)
training <- data_by_row[inTrain,]
testing <- data_by_row[-inTrain,]
```

## Model Selection

Various classification training methods, such as randomForests, ctree, ada, gbm were tested with various sets of preProcessed training data. The gbm method performed best for a selection of 10 training sets with 3% of the full data. This was subsequently selected for the final model. 

```{r eval=FALSE}
model = train(training$classe ~ ., method="gbm", data=trainPP)
```

## Model Validation

The model was then validated by reviewing the confusion matrix and calculating the accuracy on the test data set. The accuracy was in excess of 95% on the training data set. Repetitive runs using different seeds to select the training data set confirmed this as a reliable level of accuracy. 

```{r eval=FALSE}
cmatrix_train = table(train_result,training$classe)
accuracy_train = sum(diag(cmatrix_train))/sum(cmatrix_train)
accuracy_train
cmatrix_test = table(test_result,testing$classe)
cmatrix_test
accuracy_test = sum(diag(cmatrix_test))/sum(cmatrix_test)
accuracy_test
```

# Apendix - Full Final Model
```{r}
library(caret)
library(randomForest)
library(bigrf)
library(SDMTools)
library(mboost)
library(party)

# importing the data
setwd("~/docs/Private/Data Science Track/Pracitcal Machine Learning/project")

data_columns_row=c(
    "roll_belt",
    "pitch_belt",
    "yaw_belt",
    "total_accel_belt",
    "gyros_belt_x",
    "gyros_belt_y",
    "gyros_belt_z",
    "accel_belt_x",
    "accel_belt_y",
    "accel_belt_z",
    "magnet_belt_x",
    "magnet_belt_y",
    "magnet_belt_z",
    "roll_arm",
    "pitch_arm",
    "yaw_arm",
    "gyros_arm_x",
    "gyros_arm_y",
    "gyros_arm_z",
    "accel_arm_x",
    "accel_arm_y",
    "accel_arm_z",
    "magnet_arm_x",
    "magnet_arm_y",
    "magnet_arm_z",
    "roll_dumbbell",
    "pitch_dumbbell",
    "yaw_dumbbell",
    "gyros_dumbbell_x",
    "gyros_dumbbell_y",
    "gyros_dumbbell_z",
    "accel_dumbbell_x",
    "accel_dumbbell_y",
    "accel_dumbbell_z",
    "magnet_dumbbell_x",
    "magnet_dumbbell_y",
    "magnet_dumbbell_z",
    "roll_forearm",
    "pitch_forearm",
    "yaw_forearm",
    "total_accel_forearm",
    "gyros_forearm_x",
    "gyros_forearm_y",
    "gyros_forearm_z",
    "accel_forearm_x",
    "accel_forearm_y",
    "accel_forearm_z",
    "magnet_forearm_x",
    "magnet_forearm_y",
    "magnet_forearm_z",
    "classe")

data = read.csv(file="pml-training.csv", header=T, sep=",")
data_by_row = data[,data_columns_row]

proj_data =  read.csv(file="pml-testing.csv", header=T, sep=",")
proj_data_by_row = proj_data[,data_columns_row[1:50]]

set.seed(20150523)


# building training data
inTrain <- createDataPartition(y=data_by_row$classe, p=0.3, list=F)
training <- data_by_row[inTrain,]
testing <- data_by_row[-inTrain,]
project <- proj_data_by_row
dim(training)
dim(testing)


# Doing PCA
preProc <- preProcess(training[,1:50], method=c('center', 'scale'))
trainPP <- predict(preProc,training[,1:50])
testPP <- predict(preProc,testing[,1:50])
projectPP <- predict(preProc,project)

model = train(training$classe ~ ., method="gbm", data=trainPP, verbose=FALSE)

train_result = predict(model, newdata=trainPP)
test_result = predict(model, newdata=testPP)
project_result = predict(model, newdata=projectPP)


cmatrix_train = table(train_result,training$classe)
accuracy_train = sum(diag(cmatrix_train))/sum(cmatrix_train)
accuracy_train
cmatrix_test = table(test_result,testing$classe)
cmatrix_test
accuracy_test = sum(diag(cmatrix_test))/sum(cmatrix_test)
accuracy_test

project_result

```